Extreme heterogeneity of influenza virus infection in single cells

### 2018-01-16 paper/paper.tex
a few small typo / wording fixes

Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}).
However, we observe cells that lack each of the other non-RNP genes but still derive $\approx$40\% of their mRNAs from virus, suggesting that none of the other genes are important for high levels of viral transcription.
These results are statistically supported by \FIG{fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any other viral gene is associated with reduced amounts of viral mRNA. 
However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express.
Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}).  
Furthermore, the actual distribution of viral mRNA per infected cell (\FIG{viralbarcodes}E) is does not match the mostly bi-modal shape expected under a simple model where RNP gene absence and Poisson co-infection are the only factors (\FIG{fluburdenbyflugene}--source~data~\ref{figdata:simplemodel}), indicating that there are additional sources of variability beyond whether cells have full complement of RNP genes.

@@

In contrast to the extreme variability in the total viral mRNA per cell, the fraction of this mRNA derived from each gene is much more consistent across cells (\FIG{fluexpr}A).
Total viral mRNA varies by orders of magnitude, but the fraction from any given viral gene is fairly tightly clustered around the median value for all cells (\FIG{fluexpr}B).
The relative levels of each viral mRNA in our cells are similar to prior bulk measurements made by Northern blots~\citep{hatada1989control}, which also found an expression hierarchy of M $>$ NS $\gg$ NP $>$ NA $>$ HA $\gg$ PB2 $\sim$ PB1 $\sim$ PA.
The cell-to-cell consistency in the relative expression of different viral genes is even tighter if we limit the analysis only to cells that express all eight viral genes (\FIG{fluexpr}C,D).
Therefore, with the exception of complete gene absence, the factors that drive the dramatic cell-to-cell variability in the amount of viral mRNA have roughly similar effects on all viral genes in a given cell.
This finding is consistent with prior work showing positive correlations among the abundance of several viral genome segments in individual cells~\citep{Heldt:2015iz}.

@@

\begin{figure}[t!]
\centerline{\includegraphics[width=0.63\linewidth]{figures/p_cellular_heatmap_manualedits.pdf}}
\caption{\label{fig:cellulargenes}
Cellular genes that co-vary in expression with the amount of viral mRNA in cells expressing all eight viral genes.
The columns are cells, ordered from left to right by the fraction of mRNA derived from virus.
Each row is a gene that is differentially expressed as a function of the fraction of mRNA derived from virus at a false discovery rate of 0.1. 
Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA.
The scale bar indicates the number of standard deviations above or below the mean expression, truncated at 3-fold on both sides. 
}

@@

Interestingly, none of the cellular genes that are significantly associated with the amount of viral mRNA in our study are among the 128 genes that \citet{watanabe2010cellular} report as having been identified multiple times in genome-wide screens for factors affecting influenza virus replication.
One possible explanation is that most of the cell-to-cell heterogeneity in our experiments might arise from viral factors, pure stochasticity, or more subtle alterations in host-cell state -- not due to changes in expression of the type of single large-effect genes that are usually identified in genome-wide knockdown / knockout studies.

>>>>

Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}).
However, we observe cells that lack each of the other non-RNP genes but still derive $\approx$40\% of their mRNAs from virus, suggesting that none of the other genes are important for high levels of viral transcription.
These results are statistically supported by \FIG{fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any other viral gene is associated with reduced amounts of viral mRNA. 
However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express.
Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}).  
Furthermore, the actual distribution of viral mRNA per infected cell (\FIG{viralbarcodes}E) does not match the mostly bi-modal shape expected under a simple model where RNP gene absence and Poisson co-infection are the only factors (\FIG{fluburdenbyflugene}--source~data~\ref{figdata:simplemodel}), indicating that there are additional sources of variability beyond whether cells have full complement of RNP genes.

@@

In contrast to the extreme variability in the total viral mRNA per cell, the fraction of this viral mRNA derived from each gene is much more consistent across cells (\FIG{fluexpr}A).
Total viral mRNA varies by orders of magnitude, but the fraction from any given viral gene is fairly tightly clustered around the median value for all cells (\FIG{fluexpr}B).
The relative levels of each viral mRNA in our cells are similar to prior bulk measurements made by Northern blots~\citep{hatada1989control}, which also found an expression hierarchy of M $>$ NS $\gg$ NP $>$ NA $>$ HA $\gg$ PB2 $\sim$ PB1 $\sim$ PA.
The cell-to-cell consistency in the relative expression of different viral genes is even tighter if we limit the analysis only to cells that express all eight viral genes (\FIG{fluexpr}C,D).
Therefore, with the exception of complete gene absence, the factors that drive the dramatic cell-to-cell variability in the amount of viral mRNA have roughly similar effects on all viral genes in a given cell.
This finding is consistent with prior work showing positive correlations among the abundance of several viral genome segments in individual cells~\citep{Heldt:2015iz}.

@@

\begin{figure}[t!]
\centerline{\includegraphics[width=0.65\linewidth]{figures/p_cellular_heatmap_manualedits.pdf}}
\caption{\label{fig:cellulargenes}
Cellular genes that co-vary in expression with the amount of viral mRNA in cells expressing all eight viral genes.
The columns are cells, ordered from left to right by the fraction of mRNA derived from virus.
Each row is a gene that is differentially expressed as a function of the fraction of mRNA derived from virus at a false discovery rate of 0.1. 
Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA.
The scale bar indicates the number of standard deviations above or below the mean expression, truncated at 3-fold on both sides. 
}

@@

Interestingly, none of the cellular genes that are significantly associated with the amount of viral mRNA in our study are among the 128 genes that \citet{watanabe2010cellular} report as having been identified multiple times in genome-wide screens for factors affecting influenza virus replication.
One possible explanation is that most of the cell-to-cell heterogeneity in our experiments might arise from viral segment absence or mutations, pure stochasticity, or more subtle alterations in host-cell state -- not due to changes in expression of the type of single large-effect genes that are usually identified in genome-wide knockdown / knockout studies.
